# Seldonian Algorithms for Classification {#chap-3}

```{r, include = FALSE}
library(mosaic)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
```

```{r, echo = FALSE}
library(reticulate)
use_python("/cm/shared/apps/amh-Rstudio/python-3.11.4/bin/python3", required = TRUE)
```

```{r, echo = FALSE}
seldonian <- import("seldonian")
```

```{python, echo = FALSE}
import seldonian
```

Chapter \@ref(chap-2) introduced the Seldonian framework, which offers probabilistic guarantees for satisfying defined behavioral constraints. However, the toy linear regression example demonstrated some of the limitations of Seldonian algorithms, particularly in scenarios with limited sample sizes where convergence may pose a challenge. In practice, sample sizes vary. 

Furthermore, many machine learning applications deal with classification problems. Applications range from risk assessment tools like COMPAS, discussed in Chapter \@ref(intro), to credit scoring and employment prediction algorithms, naming a few. However, the deployment of such algorithms raises significant ethical concerns as discussed in Chapters \@ref(intro) and \@ref(chap-2), primarily with regard to fairness and the potential reinforcement of discriminatory practices. Given the historical and social biases inherent in the data used to train these algorithms, there is a pressing need to assess their fairness and mitigate any potential harm they may cause to disadvantaged groups.

To address this concern, this chapter aims to investigate the efficacy and applicability of Seldonian algorithms in practical classification settings. By conducting a simulation study, the aim is to evaluate whether Seldonian approaches can effectively produce fairer outcomes and mitigate discriminatory tendencies, drawing on some of the statistical definitions of fairness from Chapter \@ref(fairnessdefinitions). Specifically, the objective is to assess the feasibility of leveraging Seldonian algorithms to enhance the fairness and equity of predictive models across various classification tasks.

## Simulation Design {#sim-design}

Before conducting the simulation study and analyzing the results, this section provides detailed explanations of the simulation set-up and design. 

### Aims

### Data-Generation Mechanism

### Target

### Methods

### Performance Measures


## Simulation Results {#sim-results}


